Introduction

Please fill out this Rmarkdown file. The answers provided should look like the following:



1. Vectors

Assign a vector of 11 elements (1, 0.3, 5, 6, 5, 3, 3.5, 4, 6, 9, 9.7) to the variable test_a.
 Please provide your code here:

1.1
```{r}
test_a <- c(1, 0.3, 5, 6, 5, 3, 3.5, 4, 6, 9, 9.7)
```

Add 2 to each element of the vector test_a and assign the result to the variable test_b. Please provide your code here:

1.2
```{r}
test_b <- test_a + 2
```

Calculate the difference between test_a and test_b. Please provide your code here:

1.3
```{r}
difference <- test_a - test_b
```

Select the 3rd element of the vector test_a.
 Please provide your code here:

1.4
```{r}
test_a[3]
```

Select the 3rd, 4th and 5th elements of the vector test_a and assign the result to the vector test_c. Please provide your code here:

1.5
```{r}
test_c <- test_a[3:5]
```

Print test_c. Please provide your code here:
```{r}
test_c
```

1.6

Factors

Create the vector test_d with the elements "apple", "lemon", "orange".

2.1
```{r}
test_d <-c("apple", "lemon", "orange")
```

Convert them as factors to the variable test_e.

2.2
```{r}
test_e <- factor(test_d)
```

Please upload this Rmd file into the folder CW1 of your course repository.
